---
title: "Spatial break-down"
author: Jakub Nowosad
date: "`r Sys.Date()`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **spatialexplain** package provides model agnostic tools for exploring and explaining spatial machine learning models. 
This vignette shows a simple example of how to use it to explain a regression model with the Break Down method.

Lets start by attaching the necessary packages, reading the predictors raster, and loading the pre-trained regression model.

```{r setup}
#| message: false
# attaching the necessary packages
library(spatialexplain)
library(terra)

# reading the predictors raster
predictors = rast("/vsicurl/https://github.com/Nowosad/IIIRqueR_workshop_materials/raw/refs/heads/main/data/predictors.tif")
plot(predictors, axes = FALSE)

# loading the pre-trained regression model
data("regr_exp", package = "spatialexplain")
regr_exp
```

Now, we can use the `predict_spatial_parts()` function to explain the regression model using the Break Down method.

```{r}
regr_psp_bd = predict_spatial_parts(regr_exp, predictors, maxcell = 500,
                                    type = "break_down")
plot(regr_psp_bd)
```

The outcome is a set of raster layers that show the contribution of each predictor to the model prediction.
The Break Down method quantifies how each explanatory variable contributes to the model's average prediction by assessing the impact on the prediction as values of each variable are fixed in sequence.
Each of the cells in the raster is the contribution of the predictors to the model prediction.
Here, we can see that the intercept of the model is 15.1 for the whole area, and that the variables `popdens`, `coast`, and `ndvi` do not have any influence on the model prediction.
On the other hand, the variables `dem`, `lst_day`, and `lst_night` affect the model prediction differently, depending on the location.
For example, the `dem` variable negatively influences the model prediction in the mountainous areas -- the higher the elevation, the lower the temperature.
On the other hand, lower values of the `lst_day` and `lst_night` variables are associated with lower temperatures.

The Break Down method outcomes can be used to better understand the model's behavior and to identify the areas where the model is most sensitive to changes in the predictors.
On the other hand, it may not be the best method to explain complex models with correlated features.
For more explanation of the Break Down method, read the ["Break-down Plots for Additive Attributions"](https://ema.drwhy.ai/breakDown.html) chapter from the [Explanatory Model Analysis](https://ema.drwhy.ai/) book. 

```{r}
#| echo: false
#| eval: false
regr_psp_bdi = predict_spatial_parts(regr_exp, predictors, maxcell = 500,
                                    type = "break_down_interactions")
plot(regr_psp_bdi)
```
